---
title: "HW2_VarDistributions"
author: "Adrianna Pops"
date: "2024-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Response Variables and Suitable Distribution

Identify at least one response or outcome variable. What would be a suitable probability distribution
for your response variable?

One response variable I would like to observe is Video Like Count. Like Count is a continuous variable that I think can be modelled with a Poisson distribution.

```{r vardist}
library(tidyverse)
tiktok = read.csv("/Users/ampops/Desktop/tiktok.csv")
as_tibble(tiktok)
```
## Summary Stats

Present summary statistics of the variable(s) in your dataset using nice tables.

```{r sumstats}
tt_eda <- select(tiktok, X, claim_status, verified_status, author_ban_status, video_duration_sec, video_view_count, video_like_count)
as_tibble(tt_eda)
summary(tt_eda)
```

## Exploratory Data Analysis

Carry out an exploratory data analysis (EDA) by producing appropriate graphics to investigate
variations and co-variations in categorical and continuous variables. Be creative in presenting variations
and co-variations.

```{r graphics}
library(ggplot2)
ggplot(data = tt_eda) +
  geom_bar(mapping = aes(x = author_ban_status, fill = verified_status), position = "dodge") + 
  xlab("Author Ban Status")

ggplot(data = sample_n(tt_eda, size = 1000), aes(x = video_duration_sec, y = video_like_count, color = verified_status)) +
  geom_point() + geom_smooth() + xlab("Video Duration in Seconds") + ylab("Video Like Count") 

ggplot(data = sample_n(tt_eda, size = 1000), aes(x = author_ban_status, y = video_like_count, color = verified_status)) +
  geom_boxplot() + xlab("Author Ban Status") + ylab("Video Like Count")
```

## The Process
Based on the findings of your EDA, briefly discuss the steps of a statistical modelling process.

Using the pre-scrubbed tiktok dataset provided I first wanted to see what is the relationship if any between verified and non verified users and ban statuses for my categorical graphic. 

To visualize co-variation for continuous variables I wanted to use video duration and video like count based on verified status. 

Finally, For the combination of categorical and continuous variables I wanted to see what is the relationship if any between verified status and video like count using ban status as an explanatory variable. 

Once those were decided, I needed to determine the appropriate model to visualize the data. Because the numbers in this dataset were large I took a random sample of 1000 so the points and bars can be better seen on the graph. To fit the data to a model I would need to estimate the parameters based on the chosen distribution.  

## Feedback

Good job! It would enhance the analysis if you could provide explanations and interpretations for the graphs you've generated. Remember, the purpose of exploratory data analysis is to gain a deeper understanding of the variables, so providing insights into the patterns or trends observed in the graphs will greatly contribute to the overall analysis.


The last thing is statistical modeling process comprises the following key steps:

1. Specifying the Model:
In Step 1, the model is defined by specifying an equation for the response variable or a function thereof. This equation links the response to an optimal set of explanatory variables.

2. Estimating Model Parameters:
Moving on to Step 2, the model parameters are estimated, and the precision of these estimates is rigorously assessed.

3. Checking Model Adequacy
Step 3 involves a critical evaluation of the model's adequacy.

4. Inferences and Predictions:
Finally, in Step 4, the established model is utilized for making insightful inferences or predictions.
