---
title: "HW7_S24_Pops_Adrianna"
author: "amp-byte"
date: "2024-02-20"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 The Boston Housing Dataset

The Boston Housing Dataset is derived from information collected by the U.S. Census Service
concerning housing in the area of Boston MA. The following describes the dataset columns

+ CRIM - per capita crime rate by town

+ ZN - proportion of residential land zoned for lots over 25,000 sq.ft.

+ INDUS - proportion of non-retail business acres per town.

+ CHAS - Charles River dummy variable (1 if tract bounds river; 0 otherwise)

+ NOX - nitric oxides concentration (parts per 10 million)

+ RM - average number of rooms per dwelling

+ AGE - proportion of owner-occupied units built prior to 1940

+ DIS - weighted distances to five Boston employment centers

+ RAD - index of accessibility to radial highways

+ TAX - full-value property-tax rate per $10,000

+ PTRATIO - pupil-teacher ratio by town

+ BLACK - 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town

+ LSTAT - % lower status of the population

+ MEDV - Median value of owner-occupied homes in $1000's

**(10 pts) The objective is to predict the median value of owner-occupied homes. Identify
response variable based on the objective of the study. Run an exploratory data analysis
(EDA) to identify the relationship between the response variable and potential
predictors.**

Since we're looking to predict median value that will be our response variable. 
I'm going to measure most of the predictors against median value below. We can
find a relationship easily with the corrplot package.

```{r eda}
suppressPackageStartupMessages("tidyverse")
suppressPackageStartupMessages("glmnet")
library(tidyverse)
library(glmnet)
library(corrplot)
library(rsample)
library(ggplot2)

boston <- read.csv("/Users/ampops/Desktop/Boston.csv")
head(boston)
summary(boston)

# plotting one variable against multiple variables for univariate and bivariate 
# analysis using facet_wrap is quicker and more efficient than grid.arrange

# boxplots of predictor variables and response variable medv
boston %>%
  gather(-medv, key = "var", value = "value") %>% # dropping variable of interest
  ggplot(aes(x = ' ', y = value)) + # plotting individual variables against their values
  geom_boxplot() + # as a boxplot
  facet_wrap(~ var, scales = "free") + # from all keys and into separate panels
  theme_bw() # black and white theme

# histograms of predictor variables and response variable medv
boston %>%
  gather(-medv, key = "var", value = "value") %>% # dropping our variable of interest
  ggplot(aes(x = value)) + # plot the values of each column
  geom_histogram() + # in a histogram
  facet_wrap(~ var, scales = "free") + # from all keys and into separate panels
  theme_bw() # same theme
class(boston)
# find a relationship between response and predictor variables using color
bcor = cor(boston)
# circles varying in color to depict +/- correlation
corrplot(bcor, method = 'circle')

boston %>% # from the dataset
  gather(-medv, key = "var", value = "value") %>% # gather and convert to key value pairs medv
  ggplot(aes(x = value, y = medv)) + # plot the medv key value pairs
    geom_point() + # as a scatter plot
    stat_smooth(method = "lm", se = TRUE, col = "blue") + # with a linear regression trendline
    facet_wrap(~ var,  scales = "free") + # each in their own panel and not equal
    theme_bw() + # using a black and white theme
    ggtitle("Predictor Variables vs Response Variable medv") 

# source : Dr. Simon J Blog: https://drsimonj.svbtle.com/quick-plot-of-all-variables
```

# 2

**(10) Using the least absolute shrinkage and selection operator (LASSO) identify an
appropriate predictive model for the response variable. Write the steps including the
selection of the tuning parameter $\gamma$ for the LASSO for the Boston Housing data set.
Check if model assumptions are valid by running diagnostics on your final model. Write
the predicted model for the response variable with estimated model parameters.**

```{r lasso}
# variable selection with lasso

#split boston data 2/3 using response variable medv
splt <- initial_split(boston, prop = 0.66, strata = "medv")

# use training and test data below to run a lasso
train.dat <- training(splt) # create a training data set
test.dat <- testing(splt) # testing data set gives indexes
x_train <- model.matrix(medv~., train.dat)[, -1] # model matrix for x, -column 1
y_train <- train.dat$medv # model matrix for y

x_test <- model.matrix(medv~., test.dat)[, -1] # repeat the process for test data
y_test <- test.dat$medv # repeat the process for test data

# run several times set.seed to create reproducible results
set.seed(123)

# cross validation to get sequence of lambda values
cv.lam <- cv.glmnet(x = x_train, y = y_train, family = "gaussian", alpha = 1)
best.lam <- cv.lam$lambda.min 
# from all lambda choose small lam val where cv is small, small prediction error
# with the best lambda run the lasso again
# min lam best value for tuning parameter for cv error

x <- model.matrix(medv~., boston)[, -1]
y <- boston$medv

# shows which coefficients were shrunk to 0
# "s" specifies the tuning parameter
glm_lasso <- glmnet(x, y, family = "gaussian", alpha = 1)
plot(glm_lasso)

glm_lasso_coef <- predict(glm_lasso, type = "coefficients", s = best.lam)

glm_lasso_coef
plot(glm_lasso_coef)
```

# 3

**(10 pts) Is there any evidence of interaction between a categorical predictor and a
quantitative predictor from the EDA? If yes, run an ANCOVA for the response including
the interaction term in the final model from step 2. Check if model assumptions are
valid by running diagnostics on your model. Write the predicted model for the response
variable with estimated model parameters.**

There was a small relationship between chas and medv. 

```{r ancova}
# modelling
glm1 <- glm(medv ~ ., data = boston, family = "gaussian")
summary(glm1) # age, indus, and X are the only values not to be considered
car::vif(glm1)

par(mfrow = c(2, 2)) # mapping plots in a 2 x 2
plot(glm1)

glm2 <- glm(medv ~ crim + zn + chas + nox + rm + dis + 
              rad + tax + ptratio + black + lstat, data = boston, family = "gaussian")
summary(glm2)
par(mfrow = c(2, 2))
plot(glm2)

glm3 <- glm(medv ~ nox + rm + dis + rad + ptratio + lstat, data = boston, 
            family = "gaussian")
summary(glm3)
par(mfrow = c(2, 2))
plot(glm3)

glm4 <- glm(medv ~ nox + rm + dis + ptratio + lstat, data = boston, 
            family = "gaussian")
summary(glm4)
par(mfrow = c(2, 2))
plot(glm4)

anova(glm4, glm3, glm2, glm1) # i will use the smaller model since dev doesn't differ greatly

bosanova <- select(boston, medv, chas)
glm_bosanova<- glm(medv ~ chas, data = bosanova, family = "gaussian")
summary(glm_bosanova)

# run an ancova
glm_bosancova<- glm(medv ~ chas, data = bosanova, family = "gaussian")
summary(glm_bosancova)

anova(glm_bosancova)
```

# 4

**(5 pts) Interpret results you obtain from the models in 2) and 3).**

Based on the residuals plot I think they are all fairly similar except for the 
Q-Q plot. As I narrowed down my models I noticed the the points were following
the line marginally closer. My best Q - Q plot is in GLM Model 4, which I consider
to be my final model. There were no specific patterns in my Residuals vs Fitted
or in my Scale vs. Location plots. My Residuals vs. Leverage plots changed as I
made my way down the models. Model 1 shows it following the dotted line in middle, 
however the red line transforms and follows a positive rise.
